---
title: "Sampling distributions"
author: "Stat 250: Introduction to Statistical Inference"
output:
  xaringan::moon_reader:
    css: ["default", "assets/css/my-theme.css", "assets/css/my-fonts.css", "hygge"]
    seal: false
    lib_dir: libs
    nature:
      output:
      ratio: '16:9'
      highlightStyle: solarized-light
      highlightLanguage: ["r", "css", "yaml"]
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(dev = 'svg')
library(ggformula)
library(openintro)
library(patchwork)
library(fivethirtyeight)
library(dplyr)
library(tidyr)
library(fontawesome)

library(gridExtra)
library(tidyverse)
library(knitr)
library(mosaic)
library(kableExtra)
library(readxl)

xaringanExtra::use_panelset()

xaringanExtra::use_tachyons()

library(Sleuth3)
motivation <- case0101

brake <- read_excel("../data/brake-reaction-time.xlsx")
```

class: title-slide, left, middle
# `r rmarkdown::metadata$title`

#### `r rmarkdown::metadata$subtitle`

### `r rmarkdown::metadata$author`

---

## Roadmap

Inferential statistics relies on understanding the distribution of the statistic of interest

1. Permutation tests: estimated the distribution of the test statistic by making H<sub>0</sub> true and then using permutation resampling (Chapter 3)

2. Estimation: estimate the distribution of a statistic without assuming a specific parameter value

  - Appeal to probability (Chapter 4)
  
  - Simulate the sampling process from the population (Chapter 4)
  
  - Simulate the process using *bootstrap resampling* (Chapter 5)
  
  
---

## Example

Let $T = \displaystyle \sum_{i=1}^n X_i$ where $X_i \overset{\rm iid}{\sim} {\rm Exp}(\lambda)$. .hidden[XXX]Find the sampling distribution of $T$.

Recall: 

- PDF of $X \sim {\rm Exp}(\lambda)$: $f(x) = \lambda e^{-\lambda x}, x\ge 0$

- Moment generating function: $M_X(t) = \dfrac{\lambda}{\lambda - t}$


---

## Example

Let $X_1, X_2, \ldots, X_n$ be i.i.d. ${\rm Exp}(\lambda)$ with PDF $f(x) = \lambda e^{-\lambda}, \lambda>0, x>0$.

Find the PDF of $X_{\rm min} = \min(X_1, X_2, \ldots, X_n)$.


.footnote[Exercise 4.30]


---

## Simulation example

Let $T = \text{median} \{X_1, X_2, \ldots, X_{10} \}$ where $X_i \overset{iid}{\sim} \text{Beta}(5,2)$.

.pull-left[
.bold[Beta PDF]

$f(x) = \dfrac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)} x^{\alpha-1} (1-x)^{\beta-1}$

$0 \le x \le 1; \alpha >0, \beta >0$
]

.pull-right[
.hidden[Beta PDF]
```{r, fig.height = 2.5, fig.width = 4, out.width="100%", echo=FALSE}
gf_function(dbeta, args = list(shape1 = 5, shape2 = 2), xlim = c(0, 1)) +
  labs(y = "f(x)") +
  theme_classic()
```
]


---

## Simulation example

- Draw 10,000 samples of size $n=10$ from Beta(5, 2)

- Calculate the median for each simulated sample

```{r echo=FALSE, warning=FALSE, fig.height = 2.5, fig.width = 4, out.width="55%", fig.align='center'}
set.seed(492017) # for reproducibility

draws <- replicate(n = 10000, 
                   rbeta(n = 10, shape1 = 5, shape2 = 2))

medians <- apply(draws, 2, median)

gf_dhistogram(~medians) %>%
  gf_dens(~medians, color = "blue", size = 1) +
  theme_classic()
```

---

## Simulation example

```{r}
# Set seed for reproducibility
set.seed(492017)

# Draw 10,000 samples of size 10 from Beta(5, 2)
draws <- replicate(n = 10000, 
                   rbeta(n = 10, shape1 = 5, shape2 = 2))
# The result is a 10 x 10000 data frame 

# Calculate the median for each column
medians <- apply(draws, 2, median)
```

---

## Simulation example

We can calculate the mean, standard error, and probabilities from simulations

```{r collapse=FALSE, collapse=TRUE}
# Mean of sampling distribution
mean(medians)

# Standard error of sampling distribution
sd(medians)

# Probability of an event from sampling distribution
mean(medians <= 0.6)
```

---

## Central Limit Theorem


Let $X_1, \ldots, X_n$ be i.i.d. samples from a $\chi^2_3$ distribution.

```{r include=FALSE, cache=TRUE}
chisq_dsn <- data.frame(n10 = apply(replicate(n = 10000, rchisq(n = 10, df = 3)), 2, mean),
                        n20 = apply(replicate(n = 10000, rchisq(n = 20, df = 3)), 2, mean),
                        n40 = apply(replicate(n = 10000, rchisq(n = 40, df = 3)), 2, mean),
                        n80 = apply(replicate(n = 10000, rchisq(n = 80, df = 3)), 2, mean))

chisq_tidy <- gather(chisq_dsn, key = n, value = means)
chisq_tidy$n <- factor(chisq_tidy$n, labels = c("n = 10", "n = 20", "n = 40", "n = 80"))
chisq_tidy <- split(chisq_tidy, f = chisq_tidy$n)
```


.panelset.sideways[

.panel[.panel-name[Population]

```{r pop, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"}

x <- rchisq(10000, df = 3)
gf_dhistogram(~x, bins = 80, xlab = expression(x[i])) %>%
  gf_dens(~x, color = "blue", size = 1) +
  xlim(0, 8) +
  theme_minimal() 
  
```

]

.panel[.panel-name[n = 10]
```{r n10, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"}

gf_dhistogram(~means , data = chisq_tidy[[1]], bins = 80) %>%
  gf_dens(color = "blue", size = 1) +
  xlim(0, 8) +
  theme_minimal() 
  
```
]

.panel[.panel-name[n = 20]
```{r n20, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"} 

gf_dhistogram(~means , data = chisq_tidy[[2]], bins = 80)  %>%
  gf_dens(color = "blue", size = 1) +
    xlim(0, 8) +
  theme_minimal() 

```
]

.panel[.panel-name[n = 40]
```{r n40, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"} 

gf_dhistogram(~means , data = chisq_tidy[[3]], bins = 80)  %>%
  gf_dens(color = "blue", size = 1) +
    xlim(0, 8) +
  theme_minimal() 

```
]

.panel[.panel-name[n = 80]
```{r n80, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"} 

gf_dhistogram(~means , data = chisq_tidy[[4]], bins = 80)  %>%
  gf_dens(color = "blue", size = 1) +
    xlim(0, 8) +
  theme_minimal() 

```
]
]

---

## Continuity correction

Let $X_1, \ldots, X_n$ be i.i.d. samples from a ${\rm Bin}(n=10, p=0.75)$ distribution.


```{r include=FALSE, fig.width=5, fig.height=7}
binom_dsn <- data.frame(n10 = apply(replicate(n = 10000, rbinom(n = 10, size = 10, p = 0.75)), 2, mean),
                        n20 = apply(replicate(n = 10000, rbinom(n = 20, size = 10, p = 0.75)), 2, mean),
                        n40 = apply(replicate(n = 10000, rbinom(n = 40, size = 10, p = 0.75)), 2, mean),
                        n80 = apply(replicate(n = 10000, rbinom(n = 80, size = 10, p = 0.75)), 2, mean))


binom_tidy <- gather(binom_dsn, key = n, value = means)
binom_tidy$n <- factor(binom_tidy$n, labels = c("n = 10", "n = 20", "n = 40", "n = 80"))
binom_tidy <- split(binom_tidy, f = binom_tidy$n)
```


.panelset.sideways[

.panel[.panel-name[Population]

```{r pop2, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"}

x2 <- rbinom(10000, size = 10, p = 0.75)
gf_histogram(~x2, binwidth = .05, xlab = expression(x[i])) +
  # gf_dens(~x2, color = "blue", size = 1) +
  theme_minimal() 
  
```

]

.panel[.panel-name[n = 10]
```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"}

gf_dhistogram(~means , data = binom_tidy[[1]],  binwidth = .05) %>%
  gf_dens(color = "blue", size = 1) +
  xlim(5.5, 9.5) +
  theme_minimal() 
  
```
]

.panel[.panel-name[n = 20]
```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"} 

gf_dhistogram(~means , data = binom_tidy[[2]],  binwidth = .05) %>%
  gf_dens(color = "blue", size = 1) +
  xlim(5.5, 9.5) +
  theme_minimal() 

```
]

.panel[.panel-name[n = 40]
```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"} 

gf_dhistogram(~means , data = binom_tidy[[3]],  binwidth = .05) %>%
  gf_dens(color = "blue", size = 1) +
  xlim(5.5, 9.5) +
  theme_minimal() 

```
]

.panel[.panel-name[n = 80]
```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.height = 3.5, fig.width = 4, fig.align='center', out.width = "100%"} 

gf_dhistogram(~means , data = binom_tidy[[4]],  binwidth = .05) %>%
  gf_dens(color = "blue", size = 1) +
  xlim(5.5, 9.5) +
  theme_minimal() 

```
]
]