---
title: "The bootstrap and matched pairs"
author: "Stat 250: Introduction to Statistical Inference"
output:
  xaringan::moon_reader:
    css: ["default", "assets/css/my-theme.css", "assets/css/my-fonts.css", "hygge"]
    seal: false
    lib_dir: libs
    nature:
      output:
      ratio: '16:9'
      highlightStyle: solarized-light
      highlightLanguage: ["r", "css", "yaml"]
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "svg")
options(scipen=1, digits=3)
library(mosaic)
library(tidyverse)
library(kableExtra)
library(fontawesome)
library(readxl)
brake <- read_excel("../data/brake-reaction-time.xlsx")
```


class: title-slide, left, middle
# `r rmarkdown::metadata$title`

#### `r rmarkdown::metadata$subtitle`

### `r rmarkdown::metadata$author`

---

class: your-turn

# Your turn 1

Which of the following can be estimated from the bootstrap distribution?

(a) The mean of the sampling distribution.

(b) The standard deviation of the sampling distribution.

(c) The bias of the sampling distribution.

(d) The overall shape of the sampling distribution.

---
class: your-turn

# Your turn 2

Suppose that you have a bootstrap distribution in hand. 

What quantiles of the bootstrap distribution do you need to calculate to construct an 89% bootstrap percentile confidence interval?

---

## Is it safe to look at social media while driving?

- Previous research on smart phone use while driving has primarily focused on phone calls and texting. 

- Study looked at the effects of different smart phone tasks on car-following performance in a driving simulator. 

- Drivers performed driving only baseline simulation

- Drivers performed other phone tasks: texting, reading Facebook posts, exchanging photos on Snapchat, viewing updates on Instagram

- Brake reaction times (in seconds) recorded


.footnote[McNabb & Gray, “Staying Connected on the Road: A Comparison of Different Types of Smart Phone Use in a Driving Simulator,” *PLoS ONE*, 2016]

---

.pull-left[
```{r echo=FALSE, results='asis'}
brt <- brake %>% select(Subject, Baseline, SnapChat) %>%
  mutate(Diff = SnapChat - Baseline)
  
brt %>%
  kable(format = "html", digits = 3) %>%
  column_spec(4, background = "navy", color = "white", bold = TRUE) 
```
]

.pull-right[
## Matched pairs

- Two measurements on each case

- Goal is to estimate the true mean difference in reaction time

- Calculate differences, then it's a one-sample bootstrap

]

---

## Bootstrap percentile interval

```{r echo=FALSE, fig.height = 2.5, fig.width = 5.5, out.width = "70%", fig.align='center'}
N <- 1e4
n <- length(brt$Diff)
mean_diffs <- numeric(N)
set.seed(1112)
for(i in 1:N) {
  mean_diffs[i] <- mean(sample(brt$Diff, n, replace = TRUE))
}

quants <- quantile(mean_diffs, probs = c(0.045, 0.955))

gf_histogram(~mean_diffs, color = "gray30",
             xlab = "Mean differences", 
             title = "Bootstrap distribution",
             ylab = "Count") +
  geom_vline(xintercept = quants[1], 
             color = "darkorange") +
  geom_vline(xintercept = quants[2],
             color = "darkorange") +
  geom_label(aes(x = quants[1], y = 1000, label = round(quants[1], 3))) +
  geom_label(aes(x = quants[2], y = 1000, label = round(quants[2], 3))) +
  theme_minimal()
```

We are 89% confident that the mean difference in reaction times is between `r round(quants[1], 3)` and `r round(quants[2], 3)` seconds.

---

## Alternative resampling scheme

.left-narrow[
```{r echo=FALSE, results='asis'}
brake %>% 
  select(Subject, Baseline, SnapChat) %>%
  kable(format = "html", digits = 3) 
```
]

.right-wide[

1. Draw a **resample** of size *n* of the **rows** of the data frame, with replacement

1. Calculate the statistics on each column (e.g., means)

1. Calculate difference (or ratio, etc.)

1. Repeat this resampling process (steps 1-2) many times, say 10,000. 

1. Construct the bootstrap distribution of the statistic. 


]

---

## Bootstrap percentile interval


```{r echo=FALSE, fig.height = 2.5, fig.width = 5.5, out.width = "70%", fig.align='center'}
N <- 1e4
n <- nrow(brake)
diff_means <- numeric(N)
set.seed(1112)
for(i in 1:N) {
  df <- sample_n(brake, size = n, replace = TRUE)
  diff_means[i] <- mean(df$SnapChat) - mean(df$Baseline)
}

quants2 <- quantile(diff_means, probs = c(0.045, 0.955))

gf_histogram(~diff_means, color = "gray30",
             xlab = "Mean differences", 
             title = "Bootstrap distribution",
             ylab = "Count") +
  geom_vline(xintercept = quants2[1], 
             color = "darkorange") +
  geom_vline(xintercept = quants2[2],
             color = "darkorange") +
  geom_label(aes(x = quants2[1], y = 1000, label = round(quants2[1], 3))) +
  geom_label(aes(x = quants2[2], y = 1000, label = round(quants2[2], 3))) +
  theme_minimal()
```

We are 89% confident that the mean reaction time is between `r abs(round(quants2[1], 3))` faster to `r round(quants2[2], 3)` seconds slower when checking ShapChat compared to the baseline group.

---

class: your-turn

# Your turn

Suppose you draw two random samples from a population—one of size 50 and one of size 500—and the two sample means are equal.

You bootstrap the sample mean in both cases, using 10,000 bootstrap resamples.

How will the bootstrap distributions be similar? How will they differ?

---

class: your-turn

# Your turn

Suppose you draw two random samples from a population, both of size 50, and the two sample means are equal.

You bootstrap the sample mean in both cases.

For one bootstrap distribution, you use 1,000 bootstrap resamples, for the other you use 20,000 bootstrap resamples.

How will the bootstrap distributions be similar? How will they differ?