---
title: "Maximum likelihood estimation"
subtitle: "Stat 250"
author: "[Click here for PDF version](pdfs/09-mle.pdf)"
format:
  revealjs: 
    theme: [default, slide-styles.scss]
code-line-numbers: false
code-annotations: hover
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "svg")
options(scipen=1, digits=3)
library(tidyverse)
library(ggformula)
library(fontawesome)
```


## Joint probability, with a twist

$P(X_1 = 3, X_2 = 5, X_3 = 7) = (1-p)^{12} p^3, \quad 0 < p < 1$

```{r echo=FALSE, fig.height = 2.5, fig.width=4, fig.align='center', out.width="60%"}
gf_fun((1-x)^12 * x^3 ~ x, xlim = c(0, 1)) +
  # gf_vline(xintercept = 1/5) +
  labs(x = "p", y = "Joint probability") +
  theme_classic()
```


## Your turn

1. Let $Y_1, \ldots, Y_n$ be i.i.d. draws from a Poisson($\lambda$) distribution. Find the maximum likelihood estimator for $\lambda$.

2. Let $X_1, \ldots, X_n$ be a random sample (that is, i.i.d. draws) from a distribution with PDF 
$$f(x | \theta) = (\theta + 1) x^{\theta}, \quad 0 \le x \le 1.$$
    Find the maximum likelihood estimator for a random sample of size $n$.

## Problematic case

::::{.columns}
:::{.column width="60%"}

Suppose $X_1 = X_2 = X_3 = 1$, then $L(p) = p^3,\ 0<p<1$.

<br>

```{r echo=FALSE, fig.height = 2.5, fig.width=4, fig.align='center', out.width="90%"}
gf_fun(x^3 ~ x, xlim = c(0, 1)) +
  # gf_vline(xintercept = 1/5) +
  labs(x = "p", y = "L(p)") +
  theme_classic()
```
:::
:::{.column width="40%"}
$L(p)$ is a big as possible when $p$ is as large as possible

But $p < 1$, so $p$ can get arbitrarily close to 1, but can't reach it

$\Longrightarrow$ MLE does not exist!
:::
::::



