---
title: "Properties of Estimators"
subtitle: "Stat 250"
author: "[Click here for PDF version](pdfs/10-mle-properties.pdf)"
format:
  revealjs: 
    theme: [default, slide-styles.scss]
code-line-numbers: false
code-annotations: hover
---

```{r setup, include=FALSE}
options(scipen=1, digits=3)
library(mosaic)
library(tidyverse)
```


## {background-image="https://images.prismic.io/sketchplanations/5e043464-5a79-4ab4-b55c-5f8841e36b14_178581407694.jpg" background-size="contain"}


## Consistency illustration

Let $Y_1, Y_2, \ldots, Y_n \overset{\text{iid}}{\sim} {\text{Unif}}(\theta, \theta + 1)$. Let $\widehat{\theta} = \bar{Y}-1/2$.

```{r include=FALSE}
set.seed(123456)
n <- 1e4
theta <- 2
ys <- runif(n, theta, theta + 1)
yhats <- cumsum(ys) / seq_along(ys) - 1/2

library(ggplot2)
```

```{r echo=FALSE, fig.height = 2.5, fig.width=4.5, fig.align='center', out.width = "70%"}
ggplot(aes(y = yhats, x = seq_along(ys)), data= NULL) +
  geom_abline(aes(intercept = theta, slope = 0), color = "blue", linetype = 2) +
  geom_line() +
  labs(x = "n", y = expression(hat(theta)),
       title = "Estimation point by point",
       subtitle = expression(theta == 2)) +
  theme_classic()
```




## Poisson MLE

Let  $X_1, \ldots, X_n$ is a random sample from a $\text{Poisson}(\lambda)$ distribution

- $\widehat{\lambda} = \bar{X}$.

- $Var(\widehat{\lambda}) = Var(\bar{X}) = \lambda/n$

<br> 


**Question:** Does the likelihood tell us anything about the variance (uncertainty) of our estimator?

```{r echo=FALSE}
## A small simulation study
set.seed(9999)
N1 <- 9   # sample size 1
N2 <- 25  # sample size 2
N3 <- 100 # sample size 3

Y1 <- replicate(5, rpois(N1, lambda = 4))
Y2 <- replicate(5, rpois(N2, lambda = 4))
Y3 <- replicate(5, rpois(N3, lambda = 4))
```



## Log-likelihood functions

- 5 random samples from a ${\rm Pois}(\lambda)$ for each of $n=9, 25, 100$ drawn.
- MLE for each sample plotted in blue<br>
```{r echo=FALSE, dev = 'svg', fig.height = 3, fig.width = 9.5, out.width = "100%"}
logLik_pois <- function(lambda, ybar, n){
  log(lambda)*n*ybar - n*lambda
}

par(mfrow = c(1,3))
lambdas <- seq(2,7, len = 100)

M <- max(logLik_pois(lambdas, mean(Y3), N3))
       
plot(lambdas, logLik_pois(lambdas, mean(Y1), N1), type='l', 
     ylab = "Log Likelihood", xlab = expression(hat(lambda)), cex.lab =1.5, 
     ylim = c(0,M), cex.axis=1.5, cex.main=1.5, main = "N = 9, SE = 2/3")
abline(v=4, lwd=2, lty = 'dashed')
abline(v=colMeans(Y1), lwd=2, lty = 'dashed', col='deepskyblue')
legend("topright", lty=rep("dashed",2), col = c("black","deepskyblue"), legend = c("Truth", "MLE"), cex=1.25)

plot(lambdas, logLik_pois(lambdas, mean(Y2), N2), type='l', 
     ylab = "", xlab = expression(hat(lambda)), cex.lab =1.5,
     ylim = c(0,M), cex.axis=1.5, cex.main=1.5,main = "N = 25, SE = 2/5",
     yaxt='n')
abline(v=4, lwd=2, lty = 'dashed')
abline(v=colMeans(Y2), lwd=2, lty = 'dashed', col='deepskyblue')

plot(lambdas, logLik_pois(lambdas, mean(Y3), N3), type='l', 
     ylab = "", xlab = expression(hat(lambda)), cex.lab =1.5,
     ylim = c(0,M), cex.axis=1.5, cex.main=1.5,main = "N = 100, SE = 2/10",
     yaxt='n')
abline(v=4, lwd=2, lty = 'dashed')
abline(v=colMeans(Y3), lwd=2, lty = 'dashed', col='deepskyblue')

```

**Curvature of log-likelihood** function is linked to variability of estimator

