---
title: "From Estimators to Confidence Intervals"
subtitle: "Stat 250"
author: "[Click here for PDF version](pdfs/11-cis-estimators.pdf)"
format:
  revealjs: 
    theme: [default, slide-styles.scss]
code-line-numbers: false
code-annotations: hover
---

```{r setup, include=FALSE}
library(ggformula)
library(openintro)
library(patchwork)
library(fontawesome)

library(tidyverse)
library(knitr)
library(mosaic)
library(palmerpenguins)

library(countdown)
library(ggthemes)

gentoo <- filter(penguins, species == "Gentoo")
```


## Strategy

- We have an estimator, $\widehat{\theta}$, in hand

- Use $\widehat{\theta}$ to find a range of plausible values for $\theta$ with known long-run properties


$$P(\widehat{\theta}_{L} \le \theta \le \widehat{\theta}_{U}) = 1 - \alpha$$

## Your turn

### Finding normal quantiles in `r fa("r-project")`

`qnorm(p)` will calculate the p quantile of $N(0, 1)$


Find the value of `q` that is needed for the following $(1-\alpha)100\%$ normal-based CIs:

1. 90%

2. 95%

3. 97%

```{r echo=FALSE}
countdown(3)
```


## Your turn

Find a 90% confidence interval for the mean bill length of Gentoo penguins.

Assume that $\sigma = `r sd(~bill_length_mm, data = gentoo, na.rm = TRUE) %>% round(2)`$.

Sample statistics:

- $n=`r gentoo$bill_length_mm %>% na.omit() %>% length()`$ 

- $\bar{x} = `r mean(~bill_length_mm, data = gentoo, na.rm = TRUE) %>% round(2)`$



## Interpreting CIs redux

#### "Stat 101" answer

> We are $(1-\alpha)100$% confident that the true parameter of interest is between L and U


## Interpreting CIs redux


```{r, echo=FALSE, fig.height=4, fig.width=4, fig.align='center'}
set.seed(46453435)
POPULATION <- rnorm(n=10000, mean = 10, sd = 5)
# hist(POPULATION, main = "Population Distribution", xlab = "Y", cex.lab=1.5, cex.main=1.5, cex.axis=1.5)
# abline(v = 10, lwd = 2, col='red')
# mtext(side = 1, "Population Mean", at = 10, col='red',cex=1.5)
set.seed(2345)
n <- 10
SAMPLE <- sample(POPULATION, size = 10)
# par(mfrow = c(1,2))
# hist(POPULATION, main = "Population Distribution", xlab = "X", cex.lab=1.5, 
     # cex.main=1.5, cex.axis=1.5, xlim = c(-10,30))
# abline(v = 10, lwd = 2, col='red')
# mtext(side = 1, "Population Mean", at = 10, col='red', cex=1.5)

# hist(SAMPLE, main = "Empirical Distribution", xlab = "X", cex.lab=1.5, cex.main=1.5,
     # cex.axis=1.5, xlim = c(-10,30))
# abline(v = mean(SAMPLE), lwd = 2, col='red')
# mtext(side = 1, "Sample Mean", at = mean(SAMPLE), col='red', cex=1.5)
```

::::{.columns}
:::{.column width="50%"}
```{r, echo=FALSE, dev = 'svg', fig.height=4, fig.width = 4, out.width = "90%"}
set.seed(12245)
B <- 1e4
CI <- matrix(NA, nrow = B, ncol=2)
colnames(CI) <- c("lower", "upper")
for(b in 1:B){
  SAMPLE <- sample(POPULATION, size = 10)
  CI[b,] <- mean(SAMPLE) + c(-1,1) * 2 * 5/sqrt(10)
}

df <- as.data.frame(CI) %>%
  mutate(contains = lower <= 10 & upper >= 10) %>%
  group_by(contains) %>%
  slice_sample(prop = .01/2) %>%
  ungroup() %>%
  mutate(row = row_number()) %>%
  mutate(row = sample(row))

gf_vline(xintercept = ~10, linetype = 2) %>%
gf_linerange(factor(row) ~ lower + upper, data = df %>% head(30),
              color = ~factor(contains)) +
  scale_color_manual("Covers truth?", values = c("darkorange", "gray20")) +
  theme_classic() +
  theme(axis.line.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank())
```
:::
:::{.column width="50%"}
:::::{style="font-size: 80%;"}
- (L, U) is a random interval **before** data are observed

- The **process** by which the interval constructed is a random process

- $(1-\alpha)100$% is the long-run proportion of intervals that will capture the parameter

- In practice, we don't know which "type" of interval we have (good/bad)
:::::
:::
::::


## Plug-in principle

Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} N(\mu, \sigma^2)$.

- In practice both $\mu$ and $\sigma^2$ are unknown

- $\dfrac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim N(0, 1)$ 

- PROBLEM: $\bar{x} \pm z_{1-\alpha/2} \left( \dfrac{\sigma}{\sqrt{n}} \right)$

**Proposed solution:** plug in the sample standard deviation


## Estimating $\sigma$ impacts the distribution


```{r echo=FALSE, fig.height=3, fig.width = 6.5, out.width="95%", fig.align='center'}
z_known_sigma <- numeric(B)
z_unknown_sigma <- numeric(B)

n <- 10

for(i in 1:B) {
  s <- sample(POPULATION, size = n)
  z_known_sigma[i] <- (mean(s) - 10) / (5/sqrt(n))
  z_unknown_sigma[i] <- (mean(s) - 10) / (sd(s)/sqrt(n))
}

qq1 <- gf_qq(~z_known_sigma, title = "Distribution of Z, known SD") %>%
  gf_qqline() +
  theme_classic()
qq2 <-  gf_qq(~z_unknown_sigma, title = "Distribution of Z, unknown SD") %>%
  gf_qqline()  +
  theme_classic()

qq1 + qq2
```



## (Student's) t distribution

Let $T = \dfrac{Z}{\sqrt{V/df}}$ where $Z \sim N(0, 1)$, $V \sim \chi^2_{df}$, and $Z \perp V \Longrightarrow T \sim t_{df}$

```{r t-dsn, dev = 'svg', fig.height = 4, echo=FALSE, message=FALSE, fig.align='center'}
library(openintro)
data(COL)
par(mar = c(2, .1, .1, .1))
plot(c(-4, 4),
     c(0, dnorm(0)),
     type = 'n',
     axes = FALSE)
at <- seq(-4, 4, 2)
axis(1, at)
axis(1, at, rep("", length(at)), tcl = -0.1)
abline(h = 0)

COL. <- fadeColor(COL[5], c("FF", "89", "68", "4C", "33"))
COLt <- fadeColor(COL[5], c("FF", "AA", "85", "60", "45"))
DF   <- c('normal', 8, 4, 2, 1)

X <- seq(-4.5, 4.5, 0.02)
Y <- dnorm(X)
lines(X, Y, col = COL.[1])

for (i in 2:5) {
  Y <- dt(X, as.numeric(DF[i]))
  lines(X, Y, col = COL.[i])
}

legend(2.5, 0.4,
       legend = c(DF[1],
       paste('t, df = ', DF[2:5], sep = '')),
       col = COL.,
       text.col = COLt,
       lty = rep(1, 5))
```

## t distribution properties

::::{.columns}
:::{.column width="50%"}
- Symmetric around 0

- For $df=1$, mean doesn't exist (Cauchy distribution)

- For $df \ge 2$, $E(T) = E(Z) E \left(1 / \sqrt{V/n} \right) = 0$

- Heavier tails than normal distribution

- $t_{df} \to N(0, 1)$ as $df \to \infty$
:::
:::{.column width="50%"}
```{r t-dsn, dev = 'svg', fig.height = 3.5, fig.width= 6, echo=FALSE, message=FALSE, fig.align='center', out.width = "100%"}
```
:::
::::


## Your turn

### Finding t quantiles in `r fa("r-project")`

`qt(p, df)` will calculate the p quantile of $t_{\rm df}$

Find the value of `q` that is needed for the following $(1-\alpha)100\%$ normal-based CIs:

1. 90%, n = 123

2. 95%, n = 25

3. 99%, n = 34

```{r echo=FALSE}
countdown(3)
```


## Your turn

Find a 90% confidence interval for the mean bill length of Gentoo penguins.

Assume that $\sigma = `r sd(~bill_length_mm, data = gentoo, na.rm = TRUE) %>% round(2)`$.

Sample statistics:

- $n=`r gentoo$bill_length_mm %>% na.omit() %>% length()`$ 

- $\bar{x} = `r mean(~bill_length_mm, data = gentoo, na.rm = TRUE) %>% round(2)`$

- $s = `r sd(~bill_length_mm, data = gentoo, na.rm = TRUE) %>% round(2)`$


## Underlying validity conditions

We have a random sample from a normal population distribution

<br>

### Ask Yourself...

- Are the observations independent?

- Are the observations approximately normal?


## Checking conditions

::::{.columns}
:::{.column width="50%"}
- Are the penguins independent?

- Are the bill lengths approximately normal?
:::
:::{.column width="50%"}
```{r echo=FALSE, fig.height = 3, fig.width= 3, out.width = "100%", warning=FALSE}
gf_qq(~bill_length_mm, data = gentoo) %>%
  gf_qqline() +
  theme_classic()
```
:::
::::


## Robustness

If the a procedure "perform well" even if some of the assumptions under which they were developed do not hold, then they will be called **robust.**


## Simulation study

To check whether a procedure is robust, we can use simulation:

1. Simulate data from a variety of different probability distributions

2. Run the procedure (e.g., build a one-sample t-interval)

3. Compare the results of the procedure to what should have happened. 
    
    for a large number of CIs, approximately 95% of 95% CIs should capture the parameter value
    

##  {background-image="img/robust-1sample-t.png" background-size="contain"}


## Robustness one-sample $t$ 

:::::{style="font-size: 95%;"}
- If the population distribution is roughly symmetric and unimodal, then the procedure works well for sample sizes of at least 10–15 (just a rough guide)

- For skewed population distributions, the t-procedure can be substantially affected, depending on the severity of the skew and the sample size.

- t-procedures are not resistant to outliers.

- If observations are not independent, the results can be misleading.
:::