---
title: "Two-Sample Confidence Intervals"
subtitle: "Stat 250"
author: "[Click here for PDF version](pdfs/12-2sample-cis.pdf)"
format:
  revealjs: 
    theme: [default, slide-styles.scss]
code-line-numbers: false
code-annotations: hover
---

```{r setup, include=FALSE}
library(ggformula)
library(openintro)
library(patchwork)
library(fontawesome)
library(ggthemes)
library(readxl)

library(tidyverse)
library(knitr)
library(mosaic)
library(palmerpenguins)

library(countdown)
library(ggthemes)

library(gt)

gentoo <- filter(penguins, species == "Gentoo")
brake <- read_excel("../data/brake-reaction-time.xlsx")

```


## Bill lengths by species

```{r echo=FALSE, fig.height = 2.5, fig.width = 4, fig.align='center', out.width = "60%", warning=FALSE}
gf_boxplot(bill_length_mm ~ species, data = penguins, fill = ~species, ylab = "Bill length (mm)", alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none") +
  scale_fill_colorblind() 
```


```{r echo=FALSE}
favstats(bill_length_mm ~ species, data = penguins) %>% 
  gt() %>%
  fmt_number(columns = min:sd, decimals  = 2) |>
  tab_options(table.font.size = 24)
```

## Review: Pivotal quantity

A statistic that depends on the random sample, $X_1, \ldots, X_n$, and the parameter, $\theta$, but whose distribution <ins>does not</ins> depend on $\theta$.


## Your turn

Let $X_1,\ldots, X_n \overset{\text{iid}}{\sim} N( \mu_1, \sigma^2_1)$ and $Y_1,\ldots, Y_m \overset{\text{iid}}{\sim} N( \mu_2, \sigma^2_2)$. $X_i$ and $Y_j$ are independent.

1. Find $E(\bar{X}_n - \bar{Y}_m)$

2. Find $Var(\bar{X}_n - \bar{Y}_m)$

3. What distribution will does $Z$ have?

$$Z = \frac{(\bar{X}_n - \bar{Y}_m) - E(\bar{X}_n - \bar{Y}_m)}{\displaystyle  SD(\bar{X}_n - \bar{Y}_m)}$$
    
4. Is $Z$ a pivotal quantity?


## Your turn

Calculate and interpret a 90% confidence interval for the difference in mean bill length between Gentoo and Chinstrap penguins.


```{r echo=FALSE}

favstats(bill_length_mm ~ species, data = penguins) %>%   gt() %>%
  fmt_number(columns = min:sd, decimals  = 2) |>
  tab_options(table.font.size = 30)
```

## Underlying validity conditions

- Each random sample is taken from a normal population distribution

- The random samples are independent

:::{.fragment}

**Ask Yourself...**

- Are the observations independent?

- Are the samples of observations independent?

- Are the observations approximately normal?
:::


## Checking conditions

- Are the penguins independent within the species?

- Are the species independent?

- Are the bill lengths for each species approximately normal?

```{r echo=FALSE, fig.height = 2.5, fig.width= 7.75, out.width = "95%", warning=FALSE, fig.align='center'}
gf_qq(~bill_length_mm | species, data = penguins, color = ~species, alpha = 0.6) %>%
  gf_qqline() +
  theme_classic() +
  theme(legend.position = "none") +
  scale_color_colorblind()
```



## Robustness two-sample $t$

:::{style="font-size: 85%;"}
- If the population distribution is roughly symmetric and unimodal, then the procedure works well for sample sizes of at least 10–15 (just a rough guide)

- If the two population distributions have the same shape and the sample sizes are too different, skewness isn't a problem

- t-procedure is in trouble if the two population distributions have very different shapes (but would we even want to compare means then?)

- t-procedures are not resistant to outliers

- If observations are not independent, the results can be misleading
:::


# Revisiting matched pairs


## Is it safe to look at social media while driving?

:::{style="font-size: 85%;"}
- Previous research on smart phone use while driving has primarily focused on phone calls and texting. 

- Study looked at the effects of different smart phone tasks on car-following performance in a driving simulator. 

- Drivers performed driving only baseline simulation

- Drivers performed other phone tasks: texting, reading Facebook posts, exchanging photos on Snapchat, viewing updates on Instagram

- Brake reaction times (in seconds) recorded
:::

:::{.footer}
McNabb & Gray, “Staying Connected on the Road: A Comparison of Different Types of Smart Phone Use in a Driving Simulator,” *PLoS ONE*, 2016
:::


## CIs for matched pairs data

:::{style="font-size: 85%;"}
- For a matched pairs experiment, look at the differences between responses for each unit (pair)

- Compute a new variable for differences, then use a one-sample $t$ procedure
:::

```{r echo=FALSE}
brake |> head(8) |> gt() %>%
  fmt_number(columns = -1, decimals  = 3) |>
  tab_options(table.font.size = 24) |>
   tab_source_note(
    source_note = "And so on for 10 more rows..."
  )
```



## CIs for matched pairs data

Summary statistics for `Facebook - Baseline` brake reaction times


```{r echo=FALSE}
diff <- brake$Facebook - brake$Baseline
favstats(~diff) %>%  gt() %>%
  fmt_number(columns = min:sd, decimals  = 2) |>
  tab_options(table.font.size = 30)
```


Calculate a 92% confidence interval for the mean difference in reaction times.