---
title: "Other Confidence Intervals"
subtitle: "Stat 250"
author: "[Click here for PDF version](pdfs/13-other-cis.pdf)"
format:
  revealjs: 
    theme: [default, slide-styles.scss]
code-line-numbers: false
code-annotations: hover
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(dev = 'svg')
library(ggformula)
library(openintro)
library(patchwork)
library(fontawesome)
library(ggthemes)
library(readxl)
library(patchwork)

library(tidyverse)
library(knitr)
library(mosaic)
library(palmerpenguins)

library(countdown)
library(ggthemes)

library(gt)

survey <- read.csv("https://aloy.rbind.io/data/survey2022.csv")


gentoo <- filter(penguins, species == "Gentoo")
brake <- read_excel("../data/brake-reaction-time.xlsx")

```


## Your turn

Let $X \sim \text{Gamma}(2, \lambda)$. It is a fact that $2 \lambda X$ has a chi-square distribution with 4 degrees of freedom. Use this fact to find a 95% confidence interval for $\lambda$ via the pivotal method. 





## One-sided confidence intervals

- So far, all of our CIs were two-sided intervals of the form: $P(L \le \theta \le U) = 1-\alpha$

- When we only care about the lower or upper bound, then a one-sided interval is required

**Lower confidence bound**

Found via $P(L \le \theta)=1-\alpha$

**Upper confidence bound**

Found via $P(\theta \le U)=1-\alpha$



## Example 
:::{style="font-size: 88%;"}
- Do college students sleep < 8 hours per night?

- Average hours of sleep on weekdays collected for random sample of students from one college

- Find a 95% lower confidence bound for the average hours of sleep on weekdays for students at this college
:::

::::{.columns}
:::{.column width="50%"}

```{r echo=FALSE, fig.height = 2.5, fig.width = 4, message=FALSE, fig.align='center', out.width = "95%"}
data("SleepStudy", package = "Lock5Data")
ggplot(data = SleepStudy) +
  geom_histogram(mapping = aes(x = WeekdaySleep)) +
  labs(x = "Weekday sleep (in hours)")
```
:::


:::{.column width="50%"}
<br>


```{r results='asis', echo=FALSE}
gt(mosaic::favstats(~WeekdaySleep, data = SleepStudy)[,-9]) %>%
  fmt_number(columns = min:sd, decimals  = 2) |>
  tab_options(table.font.size = 22)
```
:::
::::


## Score intervals for proportions

:::{style="font-size: 90%;"}
A Gallup poll surveyed 3,731 randomly sampled US in April 2021, asking how they felt about requiring proof of COVID-19 vaccination for travel by airplane. The poll found that 57% said they would favor it.


Let $X =$ number of respondents in favor of proof of vaccine

Assume that $X \sim \text{Binom}(n,p)$

$\Longrightarrow \widehat{p} = X/n$ is an unbiased estimator of $p$

$\Longrightarrow$ For large $n$, $\widehat{p} \overset{\cdot}{\sim} N \left(p, \frac{p(1-p)}{n}\right)$
:::

## Score intervals for proportions

:::{style="font-size: 90%;"}
For large $n$, $Z = (\widehat{p}-p) / \sqrt{\frac{p(1-p)}{n}} \overset{\cdot}{\sim}N(0,1)$

<br>

Using the pivotal method: 

$P\left(-q \le \frac{\widehat{p}-p}{\sqrt{\frac{p(1-p)}{n}}} \le q \right) = 1-\alpha$, <br>

$\quad q = 1-\alpha/2$ quantile from $N(0,1)$

Now solve for $p$ (you'll need to remember your quadratic formula!)
:::

## Score intervals for proportions



$L = \dfrac{\widehat{p} + q^2/(2n) - q \sqrt{\widehat{p}(1-\widehat{p})/n + q^2/(4n^2)}}{1 + q^2/n}$

$U = \dfrac{\widehat{p} + q^2/(2n) + q \sqrt{\widehat{p}(1-\widehat{p})/n + q^2/(4n^2)}}{1 + q^2/n}$


:::{.aside}
We'll use R to implement this!
:::



## Example
:::{style="font-size: 90%;"}
- A Gallup poll surveyed 3,731 randomly sampled US in April 2021, asking how they felt about requiring proof of COVID-19 vaccination for travel by airplane. 

- The poll found that 57% (2,127 respondents) said they would favor it.

- Construct a 90% confidence interval for $p$
:::

```{r}
#| echo: true
prop.test(x = 2127, n = 3731, conf.level = 0.9)$conf
```
