---
title: "Classical Hypothesis Tests"
subtitle: "Stat 250"
author: "[Click here for PDF version](pdfs/14-classical-tests.pdf)"
format:
  revealjs: 
    theme: [default, slide-styles.scss]
code-line-numbers: false
code-annotations: hover
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(dev = 'svg')
library(ggformula)
library(openintro)
library(patchwork)
library(fontawesome)
library(ggthemes)
library(readxl)
library(patchwork)

library(tidyverse)
library(knitr)
library(mosaic)
library(palmerpenguins)

library(countdown)
library(ggthemes)

library(gt)
survey <- read.csv("https://aloy.rbind.io/data/survey2022.csv")

gentoo <- filter(penguins, species == "Gentoo")
brake <- read_excel("../data/brake-reaction-time.xlsx")

```

## Review: Logic of testing

:::{style="font-size: 90%;"}
1. Formulate two competing **hypotheses** about the population

2. Calculate a **test statistic** summarizing the relevant information to the claims

3. Look at the **behavior of the test statistic** assuming that the initial claim is true

4. **Compare** the observed test statistic to the expected behavior—i.e., determine the strength of evidence against the null

5. State a **conclusion in context**
:::

## One-sample test

**Do Carls sleep less than 8 hours per night?**

- Surveyed Stat 120 students 

- "On average, how many hours of sleep do you get on a weeknight?"


```{r echo=FALSE}
favstats(~sleep, data = survey) |>
  gt() %>%
  fmt_number(columns = min:sd, decimals  = 2) |>
  tab_options(table.font.size = 24)
```


```{r echo=FALSE, fig.height = 2.5, fig.width = 4, fig.align='center'}
gf_boxplot(~sleep, data = survey) %>%
   gf_refine(scale_y_discrete(labels = NULL)) +
  theme_minimal()
```

## One-sample problem

::::::{style="font-size: 85%;"}
::::{.columns}

:::{.column width=30%}
Assume
:::

:::{.column width=70%}
We have a random sample from a $N(\mu, \sigma^2)$ population
:::

::::

<br>

:::::{.fragment}
::::{.columns}

:::{.column width=30%}
Hypothesize
:::

:::{.column width=70%}
$H_0: \mu = \mu_0$ vs. $H_a: \mu \mathrel{\substack{<\\ \ne\\ >}} \mu_0$
:::

::::

:::::

:::::{.fragment}

::::{.columns}

:::{.column width=30%}
Test statistic
:::

:::{.column width=70%}
$T = \dfrac{\bar{X}-\mu_0}{s/\sqrt{n}}$
:::

::::
:::::

<br>

:::::{.fragment}

::::{.columns}

:::{.column width=30%}
Reference distribution
:::

:::{.column width=70%}
$T \sim t_{n-1}$ is $H_0$ is true
:::

::::
:::::

::::::

## Your turn

**Do Carls sleep less than 8 hours per night?**

Work with a neighbor to complete this hypothesis test and state a conclusion.

```{r echo=FALSE}
favstats(~sleep, data = survey) |>
  gt() %>%
  fmt_number(columns = min:sd, decimals  = 2) |>
  tab_options(table.font.size = 24)
```


```{r echo=FALSE, fig.height = 2.5, fig.width = 4, fig.align='center'}
gf_boxplot(~sleep, data = survey) %>%
   gf_refine(scale_y_discrete(labels = NULL)) +
  theme_minimal()
```

<br>

```{r}
countdown(4)
```



## t-tests in `r fa("r-project")`

```{r}
#| echo: true
t.test(~sleep, data = survey, mu = 8, alternative = "less")
```

:::{.aside}
`alternative` can also be set to `"greater"` or `"two.sided"`
:::




## Extracting CIs

```{r}
#| echo: true
t.test(~sleep, data = survey)$conf
```

Set `alternative` to `"greater"` or `"less"` for a one-sided confidence bound




## Example

::::::{style="font-size: 90%;"}
**Do Americans support a national health plan?**

- A Kaiser Family Foundation poll for a random sample of US adults in 2019 found that 79% of Democrats, 55% of Independents, and 24% of Republicans supported a generic "National Health Plan."

- There were 347 Democrats, 298 Republicans, and 617 Independents surveyed.

- A political pundit on TV claims that a majority of Independents support a National Health Plan. 

- Do these data provide strong evidence to support this type of statement? 
::::::

## One-proportion problem

::::{.columns}

:::{.column width=30%}
Assume
:::

:::{.column width=70%}
$X \sim \text{Binom}(n, p)$
:::

::::

<br>

:::::{.fragment}
::::{.columns}

:::{.column width=30%}
Hypothesize
:::

:::{.column width=70%}
$H_0: p = p_0$ vs. $H_a: p \mathrel{\substack{<\\ \ne\\ >}} p_0$
:::

::::

:::::

<br>

:::::{.fragment}

::::{.columns}

:::{.column width=30%}
Test statistic
:::

:::{.column width=70%}
The observed count
:::

::::
:::::

<br>

:::::{.fragment}

::::{.columns}

:::{.column width=30%}
Reference distribution
:::

:::{.column width=70%}
$X \sim \text{Binom}(n, p_0)$ if $H_0$ is true
:::

::::
:::::


## Your turn

- 339 out of 617 independents supported a National Health Plan

- Do these data provide strong evidence to support the claim that a majority of Independents support a National Health Plan? 

```{r}
countdown(4)
```



## Exact test for proportion in `r fa("r-project")`


```{r}
#| echo: true
binom.test(x = 339, n = 617, p = 0.5, alternative = "greater")
```


## Approximate test for _p_

::::::{style="font-size: 85%;"}
::::{.columns}

:::{.column width=30%}
CLT
:::

:::{.column width=70%}
For large enough $n$, $\widehat{p} \overset{.}{\sim} N \left(p, \dfrac{p(1-p)}{n} \right)$
:::

::::

<br>

:::::{.fragment}
::::{.columns}

:::{.column width=30%}
Approx. reference distribution
:::

:::{.column width=70%}
If $p=p_0$, then $\widehat{p} \overset{.}{\sim} N \left(p_0, \dfrac{p_0(1-p_0)}{n} \right)$
:::

::::
:::::

<br>

:::::{.fragment}
::::{.columns}

:::{.column width=30%}
Large-sample test statistic
:::

:::{.column width=70%}
$Z = \dfrac{\widehat{p}-p_0}{\sqrt{\dfrac{p_0(1-p_0)}{n}}}$
:::

::::
:::::

::::::


## Approximate test for proportion in `r fa("r-project")`

```{r}
#| echo: true
prop.test(x = 339, n = 617, p = 0.5, alternative = "greater")
```

## Is $n$ large enough?

- Many textbooks suggest $np_0 \ge 10$ and $n(1-p_0) \ge 10$

- Our textbook suggests $np_0 \ge 384$ and $n(1-p_0) \ge 384$

- Use the binomial test otherwise




## Two-sample t-test

:::{style="font-size: 85%;"}
**Is it better to hand write or type notes?**

- Student researchers randomly assigned 20 college students to the paper-based note-taking group and 20 students to the computer-based note taking group

-  All subjects showed a 12-minute video about the sun and allowed to take notes using the assigned method

- After video, notes collected, then subjects were given a 10-question quiz

- Does the note taking method impact the average score?

:::

## Two-sample t-test

::::::{style="font-size: 85%;"}
::::{.columns}

:::{.column width=30%}
Assume
:::

:::{.column width=70%}
- Both samples are iid draws from $N(\mu_i, \sigma^2_i)$ populations
- Independent groups
:::

::::



:::::{.fragment}
::::{.columns}

:::{.column width=30%}
Hypothesize
:::

:::{.column width=70%}
$H_0: \mu_1 - \mu_2 = \delta_0$ vs. $H_a: \mu_1 - \mu_2 \mathrel{\substack{<\\ \ne\\ >}} \delta_0$
:::

::::

:::::



:::::{.fragment}

::::{.columns}

:::{.column width=30%}
Test statistic
:::

:::{.column width=70%}
$T = \dfrac{\bar{X}_n-\bar{Y}_m-\delta_0}{\sqrt{\dfrac{s^2_1}{n}+\dfrac{s^2_1}{m}}}$
:::

::::
:::::




:::::{.fragment}

::::{.columns}

:::{.column width=30%}
Reference distribution
:::

:::{.column width=70%}
$T \sim t_{df}$ if $H_0$ is true, where df are the Welch-Sattherthwaite approx. d.f.
:::

::::
:::::

::::::

## Two-sample t-test in `r fa("r-project")`


```{r echo=FALSE}
notes <- read.csv("../data/notetaking.csv")
```

```{r}
#| echo: true
t.test(Score ~ Method, data = notes, alternative = "two.sided")
```


