---
title: "Errors and Power"
subtitle: "Stat 250"
author: "[Click here for PDF version](pdfs/15-errors-power.pdf)"
format:
  revealjs: 
    theme: [default, slide-styles.scss]
code-line-numbers: false
code-annotations: hover
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(ggformula)
library(tidyverse)
library(ggthemes)
yt <- 0
```

## {background-image="img/type_1_errors.png" background-size="contain"}


:::{.footer}
Image credit: Allison Horst
:::


## {background-image="img/type_2_errors.png" background-size="contain"}

:::{.footer}
Image credit: Allison Horst
:::


## Your turn `r (yt <- yt+1)` 

In a US court, the defendant is either innocent (H<sub>0</sub>)
 or guilty (H<sub>A</sub>)

(a) What does a Type 1 Error represent in this context? 
  
(b) What does a Type 2 Error represent?
  


## Your turn `r (yt <- yt+1)`

:::{style="font-size: 95%;"}
A consumer protection agency is testing a sample of cans of tomato soup from a company. The company is presumed to be making a safe product, unless the consumer protection agency finds evidence that the average level of the chemical bisphenol A (BPA) in tomato soup from this company is greater than 100 ppb (parts per billion). If there is evidence that the average BPA concentration is unsafe, they will recall all the soup and sue the company.

(a) State null and alternative hypotheses.

(b) What does a Type I error mean in this context?

(b) What does a Type II error mean in this context?
:::


## Type I error rate


$P(\text{Type I error}) = P(\text{Reject } H_0 |H_0 \text{ is true}) = \alpha$


AKA false positive rate

<br>

Where is this important?

- Decision rules/critical regions

- Sample size calculations


## Your turn `r (yt <- yt+1)`

- Let $Y= \#$ of successes out of $n=20$ iid S/F trials; thus, $Y \sim \text{Binom}(n=20,p)$.

- We wish to test the hypothesis $H_0 : p = .8$ versus the alternative, $H_a : p < .8$.

- Assume that the critical region $\{y \le 12\}$ is used.

- Find $\alpha$, the probability of making a type I error.



## Simulation  {background-image="img/type1errorsim.png" background-size="73%"}



```{r}
#| include: false
set.seed(46453)
POPULATION1 <- rchisq(n=10000, df=10)
POPULATION2 <- rchisq(n=10000, df=10)

set.seed(11488)

SAMPLE1 <- sample(POPULATION1, size = 25)
SAMPLE2 <- sample(POPULATION2, size = 25)

```

## Two-sample hypothesis test

$H_0: \mu_1 = \mu_2$ vs. $H_a: \mu_1 \ne \mu_2$

$$ t = \frac{\bar{X} - \bar{Y}}{\sqrt{\dfrac{s_x^2}{n} + \dfrac{s_y^2}{m}}} = -1.83$$

```{r, echo=TRUE}
t <- t.test(SAMPLE1, SAMPLE2) 
t$p.value
```


So if we reject when $p{\rm -value} < 0.1$, we would make a **Type I error**


## Distribution of p-values

```{r, echo=FALSE}
set.seed(4542353)
B <- 1e4
pval <- numeric(B)
for(b in 1:B){
  SAMPLE1 <- sample(POPULATION1, size = 25)
  SAMPLE2 <- sample(POPULATION2, size = 25)
  pval[b] <- t.test(SAMPLE1, SAMPLE2)$p.value
}
```

p-values are uniformly distributed under $H_0$

```{r, echo=FALSE, fig.height=3, fig.width=4, out.width = "40%", fig.align='center'}
hist(pval, main = NULL, xlab = "p-value", cex.lab=1.5, cex.main=1.5, breaks = seq(0,1,len=10), freq=FALSE, col="gray")
abline(v=0.1, col='darkorange', lwd=3)
arrows(0.1,0.5,0,0.5,col='darkorange', 
      length = 0.075,lwd=3)
```

True $H_0 \Longrightarrow$ 10% of the p-values are less than 0.1


## Type I error

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 2

curve(dnorm(x), lwd=2, xlim = c(-3.5,4.5), xlab = "t", 
      ylab = "Density", 
      main="", axes=F)
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)
axis(1)
axis(2)

cord.x <- c(mu,seq(mu,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(mu,4.5,len=100)),0) 

cord.x2 <- c(-mu,seq(-mu, -4.5,len=100),-3.5)
cord.y2 <- c(0,dnorm(seq(-mu,-4.5,len=100)),0) 

polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))
polygon(cord.x2,cord.y2,col=rgb(0, 0, 0,0.5))


text(x=0, y = 0.1,"Type I Error Rate")
arrows(x0=0,y0=0.075,x1=2.25,y1=0.01,length=0.1)
arrows(x0=0,y0=0.075,x1=-2.25,y1=0.01,length=0.1)
```





## Type II error rate

$P(\text{Type II error}) = P(\text{Do not reject } H_0 |H_0 \text{ is false}) = \beta$

AKA false negative rate

<br>

Where is this important?

- Power

- Sample size calculations




## Your turn `r (yt <- yt+1)`

<!-- 10.2 -->

- Let $Y= \#$ of successes out of $n=20$ iid S/F trials; thus, $Y \sim \text{Binom}(n=20,p)$.

- We wish to test the hypothesis $H_0 : p = .8$ versus the alternative, $H_a : p < .8$.

- Assume that the critical region $\{y \le 12\}$ is used and that the truth is that $p=0.6$ (so $H_0$ is false).

- Calculate $\beta$.



## Type I and type II errors

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x), lwd=2, xlim = c(-3.5,4.5), xlab = "t", 
      ylab = "Density", 
      main = "", axes=F)
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,4.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = 1, expression(paste("H"["a"]," Distribution")), col='darkorange')

cord.x <- c(2,seq(2,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(2,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(-3.5,seq(-3.5,2,len=100),2)
cord.y <- c(0,dnorm(seq(-3.5,2,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.25,y1=0.01,length=0.1)

text(x=-1.25, y = 0.3,"Type II Error Rate", col='darkorange')
arrows(x0=-1.25,y0=0.27,x1=1,y1=0.2,length=0.1, col='darkorange')
```



## Type I and type II errors

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x), lwd=2, xlim = c(-3.5,4.5), xlab = "t", 
      ylab = "Density", 
      main = "", axes=F)
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,4.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = 1, expression(paste("H"["a"]," Distribution")), col='darkorange')

cord.x <- c(2.5,seq(2.5,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(2.5,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(-3.5,seq(-3.5,2.5,len=100),2.5)
cord.y <- c(0,dnorm(seq(-3.5,2.5,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.75,y1=0.01,length=0.1)

text(x=-1.25, y = 0.3,"Type II Error Rate", col='darkorange')
arrows(x0=-1.25,y0=0.27,x1=1,y1=0.2,length=0.1, col='darkorange')
```

<!-- Smaller Type I error rate $\implies$ bigger Type II error rate -->

<!-- We decreased $\alpha$ . How did the type II error rate change? -->


## Effect size and type II errors

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x), lwd=2, xlim = c(-3.5,6.5), xlab = "t", 
      ylab = "Density", 
      main = "", axes=F)
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,6.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = 1, expression(paste("H"["a"]," Distribution")), col='darkorange')

cord.x <- c(2,seq(2,4.5,len=100),6.5)
cord.y <- c(0,dnorm(seq(2,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(-3.5,seq(-3.5,2,len=100),2)
cord.y <- c(0,dnorm(seq(-3.5,2,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.25,y1=0.01,length=0.1)

text(x=-1.25, y = 0.3,"Type II Error Rate", col='darkorange')
arrows(x0=-1.25,y0=0.27,x1=1,y1=0.2,length=0.1, col='darkorange')
```



## Effect size and type II errors

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 3

curve(dnorm(x), lwd=2, xlim = c(-3.5,6.5), xlab = "t", 
      ylab = "Density", 
      main = "", axes=F)
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,6.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = mu, expression(paste("H"["a"]," Distribution")), col='darkorange')

cord.x <- c(2,seq(2,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(2,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(-3.5,seq(-3.5,2,len=100),2)
cord.y <- c(0,dnorm(seq(-3.5,2,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.25,y1=0.01,length=0.1)

text(x=-1, y = 0.2,"Type II Error Rate", col='darkorange')
arrows(x0=-1,y0=0.17,x1=1.25,y1=0.05,length=0.1, col='darkorange')
```

<!-- The effect size increased. How did the type II error rate change? -->



## SE and type II error rate

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x), lwd=2, xlim = c(-3.5,4.5), xlab = "t", 
      ylab = "Density", 
      main = "", axes=F)
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,4.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = 1, expression(paste("H"["a"]," Distribution")), col='darkorange')

cord.x <- c(2,seq(2,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(2,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(-3.5,seq(-3.5,2,len=100),2)
cord.y <- c(0,dnorm(seq(-3.5,2,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.25,y1=0.01,length=0.1)

text(x=-1.25, y = 0.3,"Type II Error Rate", col='darkorange')
arrows(x0=-1.25,y0=0.27,x1=1,y1=0.2,length=0.1, col='darkorange')
```



## SE and type II error rate

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x, sd=0.25), lwd=2, xlim = c(-3.5,4.5), xlab = "t", 
      ylab = "Density", 
      main = "", axes=F, n=1e3)
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

curve(dnorm(x, mean=mu, sd=0.25), lwd=2, xlim = c(-3.5,4.5),
      add=TRUE, col='darkorange', n=1e3)
axis(1)
axis(2)
mtext(at = 1, expression(paste("H"["a"]," Distribution")), col='darkorange')

cord.x <- c(0.49,seq(0.49,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(0.49,4.5,len=100), sd=0.25),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(-3.5,seq(-3.5,0.49,len=100),0.49)
cord.y <- c(0,dnorm(seq(-3.5,0.49,len=100), mean=mu, sd=0.25),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=2.3, y = 0.1,"Type I Error Rate")
arrows(x0=1.7,y0=0.075,x1=1,y1=0.01,length=0.1)

text(x=-1.25, y = 0.3,"Type II Error Rate", col='darkorange')
arrows(x0=-1.25,y0=0.27,x1=0.4,y1=0.15,length=0.1, col='darkorange')
```



## Power

$$
\text{Power} = P(\text{Reject } H_0 |H_0 \text{ is false}) = 1 - \beta
$$


- Statisticians often talk about the power of a test rather than type II error

- Good tests have high power


## Your turn `r (yt <- yt+1)`

- Let $Y= \#$ of successes out of $n=20$ iid S/F trials; thus, $Y \sim \text{Binom}(n=20,p)$.

- We wish to test the hypothesis $H_0 : p = .8$ versus the alternative, $H_a : p < .8$.

- Assume that the critical region $\{y \le 12\}$ is used and that the truth is that $p=0.6$ (so $H_0$ is false).

- Calculate the power of the test.


## Power and $\alpha$

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x), lwd=2, xlim = c(-3.5,6.5), xlab = "t", 
      ylab = "Density", 
      main="", axes=F)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,6.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = mu, expression(paste("H"["a"]," Distribution")), col='darkorange')
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

cord.x <- c(2,seq(2,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(2,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(2,seq(2,6.5,len=100),6.5)
cord.y <- c(0,dnorm(seq(2,6.5,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.25,y1=0.01,length=0.1)

text(x=5, y = 0.2,"Power", col='darkorange')
arrows(x0=5,y0=0.17,x1=2.5,y1=0.05,length=0.1, col='darkorange')
```

## Power and $\alpha$

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x), lwd=2, xlim = c(-3.5,6.5), xlab = "t", 
      ylab = "Density", 
      main="", axes=F)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,6.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = mu, expression(paste("H"["a"]," Distribution")), col='darkorange')
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

cutoff <- 1.5
cord.x <- c(cutoff,seq(cutoff,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(cutoff,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(cutoff,seq(cutoff,6.5,len=100),6.5)
cord.y <- c(0,dnorm(seq(cutoff,6.5,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.25,y1=0.01,length=0.1)

text(x=5, y = 0.2,"Power", col='darkorange')
arrows(x0=5,y0=0.17,x1=2.5,y1=0.05,length=0.1, col='darkorange')
```

## Power and effect size

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x), lwd=2, xlim = c(-3.5,6.5), xlab = "t", 
      ylab = "Density", 
      main="", axes=F)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,6.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = mu, expression(paste("H"["a"]," Distribution")), col='darkorange')
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

cord.x <- c(2,seq(2,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(2,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(2,seq(2,6.5,len=100),6.5)
cord.y <- c(0,dnorm(seq(2,6.5,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.25,y1=0.01,length=0.1)

text(x=5, y = 0.2,"Power", col='darkorange')
arrows(x0=5,y0=0.17,x1=2.5,y1=0.05,length=0.1, col='darkorange')
```


## Power and effect size

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 3

curve(dnorm(x), lwd=2, xlim = c(-3.5,6.5), xlab = "t", 
      ylab = "Density", 
      main="", axes=F)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,6.5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = mu, expression(paste("H"["a"]," Distribution")), col='darkorange')
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

cord.x <- c(2,seq(2,4.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(2,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(2,seq(2,6.5,len=100),6.5)
cord.y <- c(0,dnorm(seq(2,6.5,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.3, y = 0.1,"Type I Error Rate")
arrows(x0=3.3,y0=0.075,x1=2.25,y1=0.01,length=0.1)

text(x=5, y = 0.2,"Power", col='darkorange')
arrows(x0=5,y0=0.17,x1=2.5,y1=0.05,length=0.1, col='darkorange')
```



## Power and SE

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x), lwd=2, xlim = c(-3.5,5), xlab = "t", 
      ylab = "Density", 
      main="", axes=F)

curve(dnorm(x, mean=mu), lwd=2, xlim = c(-3.5,5),
      add=TRUE, col='darkorange')
axis(1)
axis(2)
mtext(at = mu, expression(paste("H"["a"]," Distribution")), col='darkorange')
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

cord.x <- c(2,seq(2,4.5,len=100),4.5)
cord.y <- c(0,dnorm(seq(2,4.5,len=100)),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(2,seq(2,6.5,len=100),4.5)
cord.y <- c(0,dnorm(seq(2,6.5,len=100), mean=mu),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=3.75, y = 0.1,"Type I\nError Rate")
arrows(x0=3.74,y0=0.07,x1=2.25,y1=0.01,length=0.1)

text(x=3.5, y = 0.2,"Power", col='darkorange')
arrows(x0=3.49,y0=0.17,x1=2.5,y1=0.05,length=0.1, col='darkorange')
```



## Power and SE

```{r, echo=FALSE, dev = 'svg', fig.height = 4.5, fig.align='center', out.width = "73%"}
mu <- 1

curve(dnorm(x, sd=0.25), lwd=2, xlim = c(-3.5,3.75), xlab = "t", 
      ylab = "Density", 
      main = "", axes=F, n=1e3)
mtext(at = 0, expression(paste("H"[0]," Distribution")), col='black', line=1)

curve(dnorm(x, mean=mu, sd=0.25), lwd=2, xlim = c(-3.5,3.75),
      add=TRUE, col='darkorange', n=1e3)
axis(1)
axis(2)
mtext(at = 1, expression(paste("H"["a"]," Distribution")), col='darkorange')

cord.x <- c(0.49,seq(0.49,4,len=100),3.5)
cord.y <- c(0,dnorm(seq(0.49,4,len=100), sd=0.25),0) 
polygon(cord.x,cord.y,col=rgb(0, 0, 0,0.5))

cord.x <- c(0.49,seq(0.49,3.5,len=100),3.5)
cord.y <- c(0,dnorm(seq(0.49,3.5,len=100), mean=mu, sd=0.25),0) 
polygon(cord.x,cord.y,col=rgb(1, 0.5, 0,0.5), border='darkorange')

text(x=2.8, y = 0.15,"Type I\nError Rate")
arrows(x0=2,y0=0.11,x1=1,y1=0.01,length=0.1)

text(x=2.51, y = 0.5,"Power", col='darkorange')
arrows(x0=2.5,y0=0.46,x1=1,y1=0.2,length=0.1, col='darkorange')
```

## Power curves

Plot of power against different alternatives for the parameter


::::{.columns}
:::{.column width="40%"}
Example: 

- $H_0 : p = .8$ vs. $H_A : p < .8$
- $Y \sim \text{Binom}(n=20,p)$
- critical region $\{y \le 12\}$

:::

:::{.column width="60%"}

<br>

```{r echo=FALSE, fig.height = 2.5, fig.width = 4, fig.align='right', out.width="90%"}
p_grid <- seq(0.001, 0.999, by = 0.01)
power <- pbinom(12, size = 20, prob = p_grid)
power10 <- pbinom(12, size = 30, prob = p_grid)
gf_line(power ~ p_grid, xlab = "p") +
  theme_classic() + 
  scale_x_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0.01, 0.02)))
```
:::
::::

## Power curves

One-sample t-test

- $H_0: \mu = 5$ vs. $H_a: \mu \ne 5$
- Assuming $s=2$

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| warning: false
#| out-width: 60%
#| fig-align: center
mu_grid <- seq(5-3*2, 5+3*2, by = 0.05)


calc_power <- function(null = 5, truth, n = 15, s = 2, alpha = 0.05) {
  quant <- qt(1 - alpha/2, df = n - 1)
  se <- s / sqrt(n)
  cr_low <- (null - quant * se - truth) / se
  cr_high <- (null + quant * se - truth) / se
  
  pt(cr_low, df = n - 1) + 1 - pt(cr_high, df = n - 1)
}

calc_power <- Vectorize(calc_power, vectorize.args = "truth")

tpower15 <- calc_power(truth = mu_grid, n = 15)
tpower30 <- calc_power(truth = mu_grid, n = 30)
tpower60 <- calc_power(truth = mu_grid, n = 60)

tpower_df <- bind_rows(
  data.frame(n = 15, power = tpower15, mu = mu_grid),
  data.frame(n = 30, power = tpower30, mu = mu_grid),
  data.frame(n = 60, power = tpower60, mu = mu_grid),
)

gf_line(power ~ mu, data = tpower_df, linetype = ~factor(n), color = ~factor(n), xlab = expression(mu)) +
  scale_color_colorblind("n") +
  scale_linetype("n") +
  scale_x_continuous(limits = c(2, 8), expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0.01, 0.02))) +
  theme_classic() +
  theme(legend.position="top") 
```

## Test planning

If planning a hypothesis test **before** data collection, you can determine the necessary sample size to achieve a specified power when using a set significance level/rejection region.

<br>

**Example:**
Let $Y_i \overset{\rm iid}{\sim} N(\mu, 5^2)$. We wish to test $H_0: \mu=7$ vs. $H_A: \mu > 7$ at the $\alpha = 0.05$ level. What is the smallest sample size such that the test has power at least .80 when $\mu = 8$?