---
title: "Chi-squared tests"
subtitle: "Stat 250"
author: "[Click here for PDF version](pdfs/17-chi-squared-test.pdf)"
format:
  revealjs: 
    theme: [default, slide-styles.scss]
code-line-numbers: false
code-annotations: hover
---


```{r setup, include=FALSE}
library(ggformula)
library(mosaic)
library(tidyr)
library(ggthemes)
library(gt)
library(janitor)
data("happy", package = "ggmosaic")
happy2018 <- filter(happy, year == 2018) %>%
  drop_na(happy, finrela)
yt <- 0
```

## Can money buy you happiness?
:::{style="font-size: 90%;"}
The General Social Survey (GSS) is a sociological survey used to collect data on demographic characteristics and attitudes of residents of the United States. We'll consider two questions:

  *  Compared with American families in general, would you say your family income is far below average, below average, average, above average, or far above average?

  * Taken all together, how would you say things are these days—would you say that you are very happy, pretty happy, or not too happy?
:::

## Can money buy you happiness?

```{r echo=FALSE, fig.height = 4.5, fig.width = 7, out.width = "90%", fig.align='center'}
gf_bar(~finrela, data = happy2018, fill = ~happy, position = "fill") +
  xlab("Family income") +
  theme_minimal() +
  scale_fill_colorblind()
```

## Happiness contingency table


How can we explore whether opinion on income and happiness are associated?


```{r echo=FALSE}
tabyl(happy2018, happy, finrela) |>
  adorn_totals(where = c("row", "col")) |>
  gt() |>
  tab_options(table.font.size = 24)
```


R: 

```{r}
#| eval: false
#| echo: true
tabyl(happy2018, happy, finrela) |> 
  adorn_totals(where = c("row", "col"))
```

:::{.footer}
`tabyl()` and `adorn_totals()` are in the `{janitor}` package
:::


## Test statistic

$H_0:$ the variables are independent

<br>

What would the contingency table look like under $H_0$?

:::{.fragment}

```{r echo=FALSE}
chisq.test(happy2018$happy, happy2018$finrela)$expect |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "happy") |>
  gt() |>
  tab_options(table.font.size = 24)
```

:::

## Test statistic

:::{style="font-size: 90%;"}
How can we compare what we observe to what would be expected under $H_0$?

**Observed:**
:::

```{r echo=FALSE}
tabyl(happy2018, happy, finrela) |>
  adorn_totals(where = c("row", "col")) |>
  gt() |>
  tab_options(table.font.size = 24)
```

:::{style="font-size: 90%;"}
**Expected:**
:::

```{r echo=FALSE}
chisq.test(happy2018$happy, happy2018$finrela)$expect |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "happy") |>
  adorn_totals(where = c("row", "col")) |>
  gt() |>
  fmt_integer(rows = 4, sep_mark = "") |>
  tab_options(table.font.size = 24)
```



## Permutation test

:::{style="font-size: 90%;"}
1. Store the data in a table: one row per observation, one column per variable. 

2. Calculate a test statistic for the original data. 

3. **Repeat** 
  
    + Randomly permute the rows in one of the columns. 
    + Calculate the test statistic for the permuted data. 

    **Until** we have enough samples 

4. Calculate the $p$-value as the fraction of times the random statistics exceed the original statistic. 
:::


## Permutation test setup

Calculate the observed test statistic
```{r}
#| echo: true
# Have mosaic package loaded
observed_table <- tally(~ happy + finrela, data = happy2018)
observed <- chisq(observed_table)
```

<br>

Remove any missing values on variables of interest
```{r}
#| echo: true
library(tidyr) # for drop_na()
happy_complete <- drop_na(happy2018, happy, finrela)

# Extract columns of interest
happy <- happy_complete$happy
finrela <- happy_complete$finrela
```

##
### Construct the permutation distribution 

```{r cache=TRUE}
#| echo: true
set.seed(55057)
N <- 10^4 - 1
result <- numeric(N)
for(i in 1:N) {
  finrela_perm <- sample(finrela)
  perm_table <- tally(~happy + finrela_perm)
  result[i] <- chisq(perm_table)
}
```


## Permutation distribution

```{r echo=FALSE, fig.height = 2.5, fig.width = 4, out.width = "70%", fig.align='center'}
gf_histogram(~result, bins = 50, xlab = "Chi-squared statistics") + 
  theme_classic()
```


## _p_-value

```{r echo=FALSE, fig.height = 2.5, fig.width = 4, out.width = "60%", fig.align='center'}
gf_histogram(~result, bins = 50, xlab = "Chi-squared statistics") %>%
  gf_vline(xintercept = ~observed, color = "darkorange") +
  theme_classic()
```

```{r collapse=TRUE}
#| echo: true
(sum(result >= observed) + 1) / (N + 1)
```

## Chi-squared distribution

:::{style="font-size: 90%;"}
A random variable follows a $\chi^2_m$ distribution if it has PDF

$$f(x|m) = \dfrac{1}{2^{m/2} \Gamma(m/2)} x^{m/2 - 1} e^{-x/2}, \ x>0.$$
:::
```{r fig.height = 4, fig.width = 5, echo=FALSE, out.width = "40%", fig.align='center'}
x <- seq(0, 20, by = 0.1)
matplot(x, cbind(dchisq(x, df = 1), dchisq(x, df = 3), dchisq(x, df = 5), dchisq(x, df = 9)), type="l", lwd =2, ylab = "f(x)", ylim = c(0, 0.6))
```

##
### Chi-squared reference distribution

```{r echo=FALSE, fig.height = 2.5, fig.width = 4, out.width = "70%", fig.align='center'}
gf_dhistogram(~result, data = data.frame(result = result), xlab = "Chi-squared statistics", bins = 50) %>%
  gf_function(
    fun = dchisq,
    args = list(df = 8),
    color = "steelblue",
    size = 1
  ) +
  theme_classic()
```


##
### Simulation vs. model-based results

<br>

**Chi-squared test**

```{r}
#| echo: true
1 - pchisq(observed, df = (3 - 1) * (5 - 1))
```

<br>

**Permutation test**

```{r include=FALSE}
options(scipen=999)
```


```{r}
#| echo: true
(sum(result >= observed) + 1) / (N + 1)
```



## Caution

The $\chi^2$ distribution provides a reasonable approximation of the null distribution **as long as the sample size is "large enough"**

Common guidelines:

- "Cochran's rule:" All of the cells have **expected counts** > 5

- All expected counts are at least 1 and no more than 20% of cells have **expected counts** < 5



Use a permutation test if the expected counts aren't large enough



## Your turn

Work through the example on climate change action by generation with your neighbors.

<br>

R code for carrying out chi-squared tests is included on the worksheet